Event History Analyses for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-21 manuscript/manuscript.tex
update ms.

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

Integrating results between hazard and conditional accuracy functions can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for early responses (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for later bins, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

Based on Figure 6 and Table 4, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240{]} on cloglog-hazard, relative to the cloglog-estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials, in bin (200,240{]} of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360{]}, (360,400{]}, and (400,440{]}. For example, in bin (320,360{]}, the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and target-triggered response (e.g., right)

@@

Towards the end of the experiment (trial 1500), both congruent and incongruent primes have positive and negative effects. Positive effects are present in bins (200,240{]} and (240,280{]}. Incongruent primes again have negative effects in bins (320,360{]}, (360,400{]}, and (400,440{]}, and congruent primes now also have negative effects in bins (360,400{]} and (400,440{]}.
These results show that the effect of prime-target congruency changes not only on the across-bin/within-trial time scale (variable period\_9), but also on the across-trial/within-experiment time scale (variable trial\_c). The fact that congruent primes generate negative effects for 80 ms (compared to no-prime trials) towards the end of the experiment, while incongruent primes generate negative effects for 120 ms throughout the experiment, strongly suggests the involvement of two separate cognitive processes.

Panis and Schmidt (2016) distinguished between response competition effects and cognitive control effects. While response competition is present in the incongruent trials throughout the experiment, Panis and Schmidt (2016) argued that a cognitive control process is also involved, such as active top-down inhibition of the prime-triggered response. In other words, people learn that the premature prime-triggered response has to be actively inhibited in order to respond to the target stimulus. This inhibitory effect becomes visible in the congruent (compared to no-prime) trials towards the end of the experiment, while it might be masked by the inhibitory effect of response competition in the incongruent trials.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. For illustration purposes, we only fitted the effects of model M4 (see Tutorial 2a).

>>>>

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for early responses (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for later bins, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

Based on Figure 6 and Table 4, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240{]} on cloglog-hazard, relative to the cloglog-hzard estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials, in bin (200,240{]} of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360{]}, (360,400{]}, and (400,440{]}. For example, in bin (320,360{]}, the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and target-triggered response (e.g., right)

@@

Towards the end of the experiment (trial 1500), both congruent and incongruent primes have positive and negative effects. Positive effects are present in bins (200,240{]} and (240,280{]}. Incongruent primes again have negative effects in bins (320,360{]}, (360,400{]}, and (400,440{]}, and congruent primes now also have negative effects in bins (360,400{]} and (400,440{]}.

These results show that the effect of prime-target congruency changes not only on the across-bin/within-trial time scale (variable period\_9), but also on the across-trial/within-experiment time scale (variable trial\_c). The fact that congruent primes generate negative effects for 80 ms (compared to no-prime trials) towards the end of the experiment, while incongruent primes generate negative effects for 120 ms throughout the experiment, strongly suggests the involvement of separate cognitive processes.

Panis and Schmidt (2016) distinguished between automatic response competition effects (bottom-up lateral inhibition between response channels), active and global inhibition effects (top-down nonselective response inhibition), and active and selective inhibition (top-down selective response inhibition). While automatic response competition can be expected to be present in the incongruent trials throughout the experiment, active and global inhibition effects might be present in both congruent and incongruent (unmasked) prime trials. In other words, people learn that the prime-triggered response is premature and that they have to temporarily slow down (increase the global response threshold) in order to allow gating of the target stimulus. This global inhibitory effect might become visible in the congruent (compared to no-prime) trials towards the end of the experiment, while it might be masked by the automatic inhibitory effect of response competition in the incongruent trials.
Interestingly, while Panis and Schmidt (2016) did not test interactions between congruency and trial number, they concluded that active (i.e., top-down) response inhibition starts around 360 ms after the onset of the second stimulus (the target stimulus in no-mask trials), which coincides with the onset of the negative effect of congruent primes observed here in trial 1500.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. For illustration purposes, we only fitted the effects of model M4 (see Tutorial 2a).
